---
title: Exercises
jupyter:
  kernelspec:
    name: ir
    display_name: R
    language: R
---


First load the environment and install the required packages:


```{r}
renv::restore(prompt = FALSE)
```

```{r}
library(posterior)
library(priorsense)
library(bayesplot)
```

# Part 1: Posterior draws objects

For this set of exercises we will use draws from the classic eight
schools model. These draws are included in both `posterior` and `ArviZ`.

The eight schools model is a meta-analysis model of standardized test
results in different schools. The values `theta[1]` to `theta[8]` are
the mean results for each school. `mu` is the population level mean,
`tau` is the standard deviation of the population distribution.

```{r}
eight_schools_draws <- example_draws("eight_schools")
```

## Understanding draws objects

Inspect the posterior object. How many chains, iterations and variables
does it contain?

You can print the draws object, or you can use the `nchains()`,
`niterations()` and `nvariables()` functions.

```{r}
# Your code here
```

## Subsetting draws

Extract only the first chain.

You can use the `subset_draws()` function, specifying the `chain`
argument.

```{r}
# Your code here
```

Extract only the first 10 iterations, from all the chains.

You can use the `subset_draws()` function, specifying the `iteration`
argument.

```{r}
# Your code here
```

## Thinning draws

Thin the draws so that only half of the draws are included. Then try
automatic thinning.

You can use the function `thin_draws()`, optionally specifying the
`thin` argument.

```{r}
# Your code here
```

## Summarising draws

Extract the variable `mu` and summarise it via mean and sd.

You can use the functions `subset_draws()`, specifying the `variable`
argument and `summarise_draws()`, specifying the functions.

```{r}
# Your code here
```

## Creating new variables

Create a new variable that is the difference between school 1
(`theta[1]` and school 2 (`theta[2]`) means. Call it `diff_1_2`. Do the
same with school 3 (`theta[3]`) and school 4 (`theta[4]`). Then
summarise these new variables with median, 0.3 and 0.7 quantiles.

You can use the functions `mutate_variables()`, `summarise_draws()` and
`quantile2()`. You can specify variable names that include `[]` with
`` ` `` e.g. `` `theta[1]` ``.

```{r}
# Your code here
```

## Marginal posteriors

Plot the marginal posteriors for the school means.

You can use the functions `mcmc_intervals()` or `mcmc_areas()`.

```{r}
# Your code here
```

## Pairs plot

Plot the variables `mu` and `tau` in an pairs plot.

You can use the functions `mcmc_pairs()`

```{r}
# Your code here
```

# Part 2: Convergence diagnostics and uncertainty

## R-hat

Calculate the R-hat for all the variables in the draws object
(`eight_schools_draws`). Which variables have high R-hat (\> 1.01)?

You can use the function `summarise_draws()` with `rhat`

```{r}
# Your code here
```

## Effective sample size (ESS)

Calculate the bulk and tail ESS for all the variables in the model.

You can use the function `summarise_draws()` with `ess_tail` and
`ess_bulk`

```{r}
# Your code here
```

## Monte Carlo standard error

Calculate the mean of each variable, and also the Monte Carlo standard
error of the mean.

You can use the functions `summarise_draws()` with `mean` and
`mcse_mean`

```{r}
# Your code here
```

Then do the same for the 0.05 and 0.95 quantiles. Think about how the
Monte Carlo standard error might influence how you report the quantiles.

You can use the functions `summarise_draws()` with `quantile2` and
`mcse_quantile`

```{r}
# Your code here
```

## Pareto diagnostics

Calculate the minimum sample size for stable estimates for each variable
in the model. Which has the highest minimum sample size?

You can use the function `summarise_draws()` with `pareto_min_ss`.

```{r}
# Your code here
```

# Part 3: Model evaluation and critique

We can generate prior predictions from the eight schools model, using
the following function.

```{r}
eight_schools_prior <- function(ndraws, mu_prior_sd = 1, tau_prior_sd = 1, sigma_obs = c(15, 10, 16, 11, 9, 11, 10, 18)) {

  J <- length(sigma_obs)

  mu <- rnorm(ndraws, 0, mu_prior_sd)
  tau <- abs(rnorm(ndraws, 0, tau_prior_sd))

  theta <- sapply(1:J, \(j) rnorm(ndraws, mu, tau))
  yrep <- sapply(1:J, \(j) rnorm(ndraws, theta[, j], sigma_obs[j]))

  prior_draws <- as_draws(yrep)
  variables(prior_draws) <- paste0("yrep[", 1:J, "]")

  return(prior_draws)
}
```

Generate 1000 prior predictive draws, and plot the distributions for
each school. Try with different `mu_prior_sd` and `tau_prior_sd` values
(e.g. 1, 10, 100).

You can use the `ppd_intervals` function.

```{r}
prior_predictive_draws <- eight_schools_prior(
  ndraws = 1000,
  mu_prior_sd = 1,
  tau_prior_sd = 1
)

# Your code here
```

## Posterior predictive checks

We can create posterior predictive draws from our posterior draws and
plot against our actual observations.

```{r}
sigma_obs <- c(15, 10, 16, 11, 9, 11, 10, 18)

eight_schools_post_pred <- eight_schools_draws |>
  mutate_variables(
    `yrep[1]` = rnorm(length(`theta[1]`), `theta[1]`, sigma_obs[1]),
    `yrep[2]` = rnorm(length(`theta[2]`), `theta[2]`, sigma_obs[2]),
    `yrep[3]` = rnorm(length(`theta[3]`), `theta[3]`, sigma_obs[3]),
    `yrep[4]` = rnorm(length(`theta[4]`), `theta[4]`, sigma_obs[4]),
    `yrep[5]` = rnorm(length(`theta[5]`), `theta[5]`, sigma_obs[5]),
    `yrep[6]` = rnorm(length(`theta[6]`), `theta[6]`, sigma_obs[6]),
    `yrep[7]` = rnorm(length(`theta[7]`), `theta[7]`, sigma_obs[7]),
    `yrep[8]` = rnorm(length(`theta[8]`), `theta[8]`, sigma_obs[8])
  ) |>
  subset_draws(variable = "yrep") |>
  as_draws_matrix()

# Or alternatively in a nicer way, using rvars

eight_schools_post_pred <- eight_schools_draws |>
  as_draws_rvars() |>
  mutate_variables(
    yrep = rvar_rng(
      rnorm,
      n = length(sigma_obs),
      mean = theta,
      sd = sigma_obs
    )
  ) |>
  subset_draws(variable = "yrep") |>
  as_draws_matrix()
```

Plot the posterior predictions on top of the actual observations. Then
use the PIT-ECDF plot.

You can use the `ppc_intervals()` and `ppc_pit_ecdf()` functions.

```{r}
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
# Your code here
```

## Sensitivity checks

Check for prior and likelihood sensitivity in the eight schools model.
First check by power-scaling all priors jointly, then select only the
“mu” and only the “tau” prior separately.

You can use `powerscale_sensitivity()` with the `prior_selection`
argument.

```{r}
eight_schools_draws_ps <- example_powerscale_model("eight_schools")$draws
# Your code here
```

Next plot sensitivity using density plots. Plot only the mu and tau
variables.

You can use `powerscale_plot_dens()`.

```{r}
# Your code here
```

# Part 4: Bringing it all together

We have provided four sets of posterior draws from hierarchical models
of observed migratory bird counts recorded between 2000 and 2020 at the
[Hanko Bird Observatory
(Halias)](https://halias.fi/tutkimus-ja-aineisto/).

For species $j$,

$y \sim \mathrm{Poisson}(\lambda_j)$

or

$y \sim \mathrm{NegativeBinomial}(\lambda_j, \phi).$

The species-specific abundances are linked through a hierarchical prior,

$\log(\lambda_j) \sim \mathrm{Normal}(\mu, \sigma)$

where $\mu$ represents the average abundance across species and $\sigma$
controls the amount of pooling between species.

Your task is to explore the posterior draws and diagnostic outputs for
the four fitted models.

As you work through the diagnostics, try to identify which model
corresponds to each of the following situations:

- Convergence issues caused by poor chain mixing (for example,
  insufficient warmup).
- Inadequate fit to the data caused by the choice of observation model,
  shown by posterior predictive checks.
- Issues caused by priors in conflict with the likelihood.
- No major issues, although there is still room for model improvement.

Use posterior summaries, convergence diagnostics, posterior predictive
checks, and sensitivity analyses to guide your investigation.

You can also look at the Stan model (`birds_per_year.stan`) and consider
how you might improve it.

```{r}
# Use the following as a starting point

# Load the draws
draws1 <- readRDS("data/draws1.rds")
draws2 <- readRDS("data/draws2.rds")
draws3 <- readRDS("data/draws3.rds")
draws4 <- readRDS("data/draws4.rds")

# Load the data fed to Stan
stan_data <- readRDS("data/stan_data.rds")

# Get the actual observations and the corresponding species
y <- stan_data$count
species <- stan_data$species # useful for grouped plots

# Create yrep matrices for posterior predictive checks
yrep1 <-  as_draws_matrix(subset_draws(draws1, "y_rep"))
yrep2 <-  as_draws_matrix(subset_draws(draws2, "y_rep"))
yrep3 <-  as_draws_matrix(subset_draws(draws3, "y_rep"))
yrep4 <-  as_draws_matrix(subset_draws(draws4, "y_rep"))
```

